---
title: "Prediction"
author: "Ian Wheeler"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, }
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, error = FALSE)
```

## Abstract

The following analysis uses random forests to predict the quality of exercise activities performed. The dataset used is the Weight Lifting
Exercise (WLE)[^1] dataset used by permission via a Creative Commons license. The findings reveal that....

## Import the Data

```{r Import the data and view its dimensions}
library(dplyr)
library(utils)
library(caret)


url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
training <- read.csv(url)

url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
testing <- read.csv(url)

dim(training); dim(testing)
```

## Clean and pre-process the data.

```{r Clean and preprocess data}
# Remove any non-numeric columns or columns with only NA values from both training and testing data
training <- training %>% select(-c(1:7))
testing <- testing %>% select(-c(1:7))
training <- training[, colSums(is.na(training))==0]
testing <- testing[, colSums(is.na(testing))==0]
trainNumeric <- sapply(training, is.numeric)
testNumeric <- sapply(testing, is.numeric)
trainFiltered <- training[, trainNumeric] ; trainFiltered$classe <- training$classe
testFiltered <- testing[, testNumeric]
```

## Split the training data up into training, validation, and testing subsets for cross-validation.

```{r Splitting Dataset}
# Split the training data set up into training, validation, and testing sets
dataPart <- createDataPartition(y = trainFiltered$classe, p = 0.7, list = FALSE)
trainPart <- trainFiltered[dataPart,]; evalPart <- trainFiltered[-dataPart,]
dataPart2 <- createDataPartition(y=trainPart$classe, p = 0.7, list = FALSE)
trainPart <- trainPart[dataPart2,]; testPart <- trainPart[-dataPart2,]

dim(trainPart); dim(evalPart); dim(testPart); dim(testFiltered)
```
## Use cross-validation to train the Random Forests Model.

```{r Use cross-validation to train Random Forests Model}
# Use cross-validation with 10 k-folds
set.seed(1066)
trnCtrl <- trainControl(method = "cv", number = 3, verboseIter = TRUE)

# Train the Random Forests model 
model_rf <- train(classe ~ ., data = trainPart, method = "rf",
                  trContrl = trnCtrl, ntree = 200)
model_rf$finalModel

# List the most important variables in the model
varImp(model_rf)
```
## Predict on Validation and Testing Sets.

```{r Predict on Validation and Testing Sets}
evalPred <- predict(model_rf, newdata = evalPart)
testPred <- predict(model_rf, newdata = testPart)
evalPart$classe <- as.factor(evalPart$classe)
testPart$classe <- as.factor(testPart$classe)
eval_cm <- confusionMatrix(evalPred, evalPart$classe)
test_cm <- confusionMatrix(testPred, testPart$classe)
eval_cm$overall[1:2]
test_cm$overall[1:2]
```
The sample error in the validation set is `1-as.numeric(eval_cm$overall[1])` and the testing subset sample error was `1-as.numeric(test_cm$overall[1])` due to over-fitting.

## Predict new values on the original testing set and table the predicted values by problem_id.

```{r Predicting new values in testing set}
# Measure accuracy of predictions against testing set
prediction <- predict(model_rf, newdata = testFiltered)
testFiltered$classe <- prediction

# Table test cases by predicted class
table(testFiltered$problem_id, testFiltered$classe)

```

[^1]: Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.